---
title: |
  A combined recurrent neural network model for cryptocurrency time series forecasting
type: 
author:
  - name: Wiem Ben Romdhane
    affil: a
    email: wiembr@gmail
  - name: Heni Boubaker
    affil: a
    email: heniboubaker@gmail.com
 
affiliation:
  - num: a
    address: |
      LaREMFIQ, IHEC Sousse, Tunisia.
 
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  Forecasting cryptocurrency time series presents a difficult challenge, primarily attributable to their inherent non-linearity and chaotic behavior. While traditional statistical methodologies have yielded notable efficacy in specific domains, such as directional market prediction and individual equity price forecasting, the advent of neural networks and recurrent neural networks has catalyzed the exploration of novel paradigms in financial time series prediction. Furthermore, contemporary research posits that the synergistic integration of statistical and machine learning techniques can yield enhanced predictive accuracy relative to their isolated application. This study, therefore, proposes a combined framework that integrates statistical features derived from financial time series with a recurrent neural networks architecture for temporal forecasting. The efficacy of this methodology was evaluated through the prediction of Bitcoin (BTC) closing prices, utilizing a suite of performance metrics. Empirical results demonstrate the superiority of our model compared to univariate statistical and machine learning models.  
keywords: |
  Forecasting; BTC; Time series; RNN; LSTM; GRU; Combined
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

# Introduction


```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

The rapid growth of cryptocurrencies over the past decade has transformed them into a significant component of the global financial system. However, the volatile and non-linear nature of cryptocurrency prices presents unique challenges for accurate forecasting. Traditional statistical methods often fall short in capturing the intricate temporal dependencies and high volatility of cryptocurrency time series. As a result, researchers have increasingly turned to deep learning models, particularly Recurrent Neural Networks (RNNs), to address these challenges [@NASIRTAFRESHI2022; @kumar23; @seable23] .

Recurrent Neural Networks (RNNs) are well-suited for time series forecasting due to their ability to model sequential data and capture temporal dependencies. Variants of RNNs, such as Long Short-Term Memory (LSTM) and Gated Recurrent Unit (GRU), have been widely adopted for cryptocurrency forecasting. These models are particularly effective in handling long-term dependencies and mitigating the vanishing gradient problem, which is common in traditional RNNs. 
For instance, Nasirtafreshi (2022) proposed an RNN-LSTM model to predict cryptocurrency prices, demonstrating its ability to outperform traditional methods in terms of accuracy.

To further enhance the predictive performance of RNNs, researchers have explored hybrid models that combine RNNs with other techniques. For example, Guo et al. (2021) developed a hybrid method that integrates a multiscale residual block with an LSTM network to forecast Bitcoin prices. This approach leverages the strengths of both components: the residual block captures multi-scale features, while the LSTM network models temporal dependencies. 

Similarly, other studies have proposed combining RNNs with convolutional layers or graph-based methods to improve the model's ability to capture spatial and temporal patterns in cryptocurrency data. The combined RNN models offer several advantages over standalone RNNs or traditional methods. By integrating complementary techniques, these models can better handle the non-stationarity and noise inherent in cryptocurrency time series. For instance, hybrid models that incorporate wavelet transforms or attention mechanisms have been shown to improve feature extraction and focus on relevant patterns in the data. Additionally, ensemble approaches that aggregate predictions from multiple RNN-based models have demonstrated enhanced robustness and accuracy in cryptocurrency forecasting. 

# Related works

Machine Learning is an artificial intelligence device that uses beyond information to predict the future. In simple words, we can expect the future price movements of cryptocurrencies to some extent by training a machine learning model using their past price data. Some recent studies have shown that machine learning based methods have many advantages of using traditional forecasting models, such as the ability to produce results that are approximately equal or identical to the actual outcome, while also improving the precision of the outcome [@hitam2021]. Decision trees, support vector machine, and neural networks are some of the different machine learning methods that can be used for this purpose. As evidenced by the authors in [@andrianto17], the inclusion of cryptocurrencies in multi-asset portfolios significantly improves portfolios in several ways. To start, it will enhance the portfolio’s minimal variance and furthermore transfers the green frontier to a higher location. 

Several research studies in the literature that using machine learning algorithms in BTC price forecasting achieve encouraging results. According to a study [@hitam2021], machine learning algorithms were applied to perform the price prediction of many currencies including BTC, ETH, LTC, XRP, and Stellar. According to the researchers, the SVM model was able to beat other machine learning models in terms of predicted values as well as accuracy. [@saad19] used a variety of variables, carefully choosing the most accurate predictors using correlation analysis. The results showed that linear regression performed better than the other approaches when SVM, linear regression, and random forest (RF) were used to these selected features. The authors also tried predicting the prices of BTC and ETH using LSTM, a specific kind of deep learning, and discovered that LSTM had the lowest prediction error for BTC. To predict the prices of nine different cryptocurrencies, [@chowdhury20] studied the application of machine learning-based ensemble methods, namely ANN, KNN, Gradient Boosted Trees and an ensemble model made up of multiple methods. The ensemble learning model had the the lowest error in the predictions. [@derbentsev21] studies the difficulties faced when forecasting short-term cryptocurrency time series using supervised machine learning (ML). The ensemble methods were then applied to the daily closing prices of Bitcoin (BTC), Ethereum (ETH), and Ripple (XRP) using historical prices and technical indicators such as moving averages as features: Random Forest (RF) and Stochastic Gradient Boosting Machine (SGBM). The results showed that ML ensemble methods held promise, with SGBM and RF yielding good accuracy on short-term predictions. [@chen2021] introduces a two-stage approach to investigate whether economic and technology determinants can accurately predict the Bitcoin exchange rate. In the first stage, artificial neural networks and random forests are employed as nonlinear feature selection methods to identify and prioritize key predictors from economic and technology factors. In the second stage, these selected predictors are integrated into a long short-term memory (LSTM) model to forecast the Bitcoin exchange rate without relying on historical exchange rate data. The results demonstrate that LSTM, utilizing economic and technology determinants, outperforms traditional models such as autoregressive integrated moving average, support vector regression, adaptive network fuzzy inference system, and LSTM models that depend on past exchange rates. This highlights that economic and technology factors are more significant for predicting Bitcoin exchange rates than historical exchange rate information. [@patel2020] introduce a hybrid prediction model for cryptocurrencies, combining LSTM and GRU networks, specifically applied to Litecoin and Monero. The results show that our approach achieves highly accurate price predictions, demonstrating its potential for broader use in forecasting the prices of various cryptocurrencies. This suggests that the model could be a valuable tool for understanding and predicting cryptocurrency market trends. [@zhang2021] propose a model that is built around three key modules designed to enhance cryptocurrency price prediction. First, the Attentive Memory module integrates a Gated Recurrent Unit (GRU) with a self-attention mechanism, allowing the model to focus on the most relevant parts of each input sequence. Second, the Channel-wise Weighting module analyzes the prices of major cryptocurrencies, learning their relationships by dynamically adjusting the importance of each sequence. Finally, the Convolution \& Pooling module captures local patterns in the data, improving the model’s ability to generalize. To test its effectiveness, the authors ran a series of experiments, which demonstrated that their model outperforms existing baseline methods in terms of prediction error, accuracy, and profitability, achieving state-of-the-art results. [@nouira2024] explore recent advancements in machine learning and deep learning techniques used for predicting cryptocurrency prices, as highlighted in highly regarded publications. By focusing on the latest research, the authors analyze how these models leverage social network data to forecast cryptocurrency trends. The findings reveal a notable shift in the field, with deep learning methods increasingly surpassing traditional machine learning approaches in cryptocurrency price prediction. The study of [@vieitez2024] develops prediction models for Ethereum (ETH) prices using Gated Recurrent Unit (GRU) and Long Short-Term Memory (LSTM) networks, alongside a Support Vector Machine (SVM) for trend classification. Sentiment analysis is incorporated to assess its impact, though it shows minimal influence on results. Two innovative knowledge-based investment strategies are designed and tested across different time periods using real market data. The findings reveal that these models can achieve a profit factor of up to 5.16 with limited trading activity, demonstrating their reliability and potential for generating returns in the current cryptocurrency market. [@lapitskaya2024] combine technical analysis tools—such as Moving Average Convergence/Divergence, Commodity Channel Index, and Relative Strength Index—with the eXtreme Gradient Boosting (XGBoost) model to predict cryptocurrency prices and returns. Using historical daily data for Bitcoin, Ether, Golem, and FUNToken, the model achieves strong accuracy and reliable performance across both training and test datasets. This approach highlights the potential of integrating traditional indicators with machine learning for effective cryptocurrency price forecasting. Recent work of [@shirwaikar2025] focuses on predicting the values of three cryptocurrencies—Litecoin, Ethereum, and Monero—using deep learning techniques, specifically Long Short-Term Memory (LSTM) and Gated Recurrent Unit (GRU). To enhance accuracy, the research incorporates the Direction Algorithm (DA), which analyzes Bitcoin’s price movements as a benchmark for other cryptocurrencies, and Change Point Detection (CPD) using Pruned Exact Linear Time (PELT) to identify sudden price shifts. Additionally, sentiment analysis of news data is performed using the Vader algorithm to further improve forecasting precision.

```{r, echo=F, comment=NA, message=FALSE, warning=FALSE, fig.align='center', out.width="85%"}
library(tidyverse, quietly =T)
btc <- crypto2::crypto_history(limit = 2,
  start_date = "2016-01-06",
  end_date = "2025-02-20",
  
)
btc_c=btc %>% dplyr::filter(symbol=="BTC")%>%dplyr::select(time_close, close)
BTC=xts::xts(btc_c$close, btc_c$time_close )
forecast::autoplot(BTC)+xlab('Time')
```


# Acknowledgement(s) {-}

An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Disclosure statement {-}


# Funding {-}

An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Notes on contributor(s) {-}


# Nomenclature/Notation {-}


# Notes {-}


# References



